---
title: "Analysing RiboSeq Data"
author: "Alexander Grentner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing RiboSeq Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

TODO write text here Gionmattia approx 4 sentences
TODO delete this line


## Installation

The `ReactomeGSA` package can be directly installed from Bioconductor:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

if (!require(ReactomeGSA))
  BiocManager::install("ReactomeGSA")


```

## Getting available methods

The Reactome Analysis System provides multiple pathway algorithms but only PADOG is available for Ribo-Seq data analysis. 
Therefore, to analyse Ribo-Seq data "PADOG" must be selected to choose a modified version for valid Pathway analysis


```{r, echo=FALSE}
library(ReactomeGSA)

my_request <- ReactomeAnalysisRequest(method = "Padog")
my_request
```

All methods for PADOG are available to perform Ribo-Seq analysis. 

```{r, echo=FALSE}
# show the parameter names for the method
padog_params <- available_methods$parameters[available_methods$name == "PADOG"][[1]]
paste0(padog_params$name, " (", padog_params$type, ", ", padog_params$default, ")")
```

## Load data

To perform Ribo-Seq Data analysis a Ribo-Seq file and a RNA-Seq file is needed. Both files need matching column names. Both files are then processed 

```{r, echo=FALSE}

```


## Set parameters

Set parameters for Ribo-Seq analysis request.

```{r, echo=FALSE}
my_request <- set_parameters(request = my_request, max_missing_values = 0.5)
my_request
```





```{r, echo=FALSE}
my_request <- add_dataset(request = my_request, 
                          expression_values = ribo_example, 
                          name = "Ribo Sequencing Data", 
                          type = "ribo_seq",
                          comparison_factor = "condition", 
                          comparison_group_1 = "treated", 
                          comparison_group_2 = "untreated"
                          )
my_request
```

